---
title: "Secondary analysis - comparing the two approaches (inter-treatment vs. treatment vs. controls)"
author: "Francisca Rodriguez"
date: "12/3/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Data input**
```{r}
#load libraries 
library(phyloseq)
library(tibble)
library(dplyr)

#Load seqtable.nochim and taxa from dada2 files from Kacie
seqtable.nochim <- readRDS('./raw_data/dada2_seqtable.RDS')
taxa <- readRDS('./raw_data/RDP_Taxonomy_from_dada2.RDS')
strawb_miseq_map <- read.csv("./raw_data/Strawberry_Miseq_Map_forQIIME.csv", fileEncoding="UTF-8-BOM")

# add empty column
strawb_miseq_map <- add_column(strawb_miseq_map, SampleType = NA)

# get rid of unneeded cols
strawb_miseq_map <- select(strawb_miseq_map, 
                           SampleID, 
                           TimePoint, 
                           Type, 
                           PlantID, 
                           WeightGrams, 
                           SampleNotes, 
                           Water, 
                           Inoculation, 
                           Replicate, 
                           PlantNotes, 
                           SampleType)

# label sample types *INOCULATION 6 INCLUDED IN POST_INOC_POST_STRESS*
strawb_miseq_map$SampleType[strawb_miseq_map$TimePoint == "T0"] <- "post_inoc_pre_stress"
strawb_miseq_map$SampleType[strawb_miseq_map$TimePoint == "T1"] <- "post_inoc_post_stress"
strawb_miseq_map$SampleType[strawb_miseq_map$SampleNotes == "Empty Extraction Tube Control"] <- "DNA_neg"
strawb_miseq_map$SampleType[grep("Inoc", strawb_miseq_map$SampleNotes)] <- "slurry"
strawb_miseq_map$SampleType[grep("Inocolum 6", strawb_miseq_map$SampleNotes)] <- "di_water"
strawb_miseq_map$SampleType[grep("pre-inoc", strawb_miseq_map$SampleNotes)] <- "pre_inoc"

# add empty column for control
strawb_miseq_map <- add_column(strawb_miseq_map, Sample_or_Control = "True Sample")

#label controls (i6, DI water, DNA/PCR negative, pre-inoculated plants)
strawb_miseq_map$Sample_or_Control[strawb_miseq_map$SampleType == "di_water"] <- "Control"
strawb_miseq_map$Sample_or_Control[strawb_miseq_map$SampleType == "DNA_neg"] <- "Control"
strawb_miseq_map$Sample_or_Control[strawb_miseq_map$SampleType == "pre_inoc"] <- "Control"
strawb_miseq_map$Sample_or_Control[strawb_miseq_map$Inoculation == "6" & strawb_miseq_map$TimePoint == "T1"] <- "Control"

# remove unwanted samples
strawb_miseq_map <- strawb_miseq_map[grep("CSIA", strawb_miseq_map$SampleNotes, invert = TRUE),]
strawb_miseq_map <- strawb_miseq_map[strawb_miseq_map$Type != "Palmyra",]

#save new map as a .csv file
write.csv(strawb_miseq_map, "./combined_treatment_data/strawb_miseq_map2.csv")
```

**Phyloseq**
```{r}
# Hand off to Phyloseq ####
otu <- otu_table(seqtable.nochim,taxa_are_rows = FALSE)
tax <- tax_table(taxa)
met <- sample_data(strawb_miseq_map)
row.names(met) <- strawb_miseq_map$SampleID

ps2 <- phyloseq(otu,met,tax)

# generate the ASV0001, ASV0002 names
pretty_names <- paste("ASV", sprintf('%0.4d', 1:length(taxa_names(ps2))), sep = "")

# reassign the taxa_names from the sequences to the generated ASV names
taxa_names(ps2) <- pretty_names

#save phyloseq object
saveRDS(ps2, "./combined_treatment_data/ps2.rds")

```

**Decontaminate controls**
```{r}
library(decontam)

#remove taxa with only one read 
ps2 <- prune_taxa(taxa_sums(ps2) > 1, ps2)

#label controls (extraction negatives) and clean
sample_data(ps2)$is.neg <- sample_data(ps2)$Sample_or_Control == "Control"
contams <- isContaminant(ps2, method="prevalence", neg="is.neg")
table(contams$contaminant)

#Remove contaminants
ps.noncontam2 <- prune_taxa(!contams$contaminant, ps2)

#save phyloseq object
saveRDS(ps.noncontam2, "./combined_treatment_data/ps.noncontam2.rds")
```

**Subset by post-inoc post-stress**
```{r}
#post_inoc_post_stress table 
post_inoc_post_stress_ps <- subset_samples(ps.noncontam2, sample_data(ps.noncontam2)$SampleType=="post_inoc_post_stress")
post_inoc_post_stress_ps <- prune_taxa(taxa_sums(post_inoc_post_stress_ps) > 0, post_inoc_post_stress_ps)
post_inoc_post_stress_asv <- as.data.frame(otu_table(post_inoc_post_stress_ps))

#Hellinger transformation for all replicates in post-inoculation/post-stress ASV contingency table
transformed_pips_ASV <- hellinger(post_inoc_post_stress_asv)

#On this transformed data, calculate a “relative abundance” by summing the new totals for all ASVs (the transformed number of total reads) within a replicate and then dividing the transformed number of reads for each ASV by this total
rel_abund <- transformed_pips_ASV / rowSums(transformed_pips_ASV) 

#pull out tax table and metadata from post_inoc_post_stress_ps to create new phyloseq object
pips_metadata <- as(sample_data(post_inoc_post_stress_ps), 'matrix')
pips_metadata <- as.data.frame(pips_metadata)
pips_taxtable <- as(tax_table(post_inoc_post_stress_ps), 'matrix')

#Create new phyloseq object with transformed table
otu <- otu_table(rel_abund,taxa_are_rows = FALSE)
tax <- tax_table(pips_taxtable)
met <- sample_data(pips_metadata)
row.names(met) <- pips_metadata$SampleID

ps.pips2 <- phyloseq(otu,met,tax)

#save phyloseq object
saveRDS(ps.pips2, "./combined_treatment_data/ps.pips2.rds")
```

**Box and whisker plot of average richness**
```{r}
#Sum ASVs across samples to add to metadata
richness_sums <- rowSums(post_inoc_post_stress_asv)

#add column to new df with corresponding richness values
pips_complete2 <- dplyr::bind_cols(pips_metadata, richness_sums)

#rename column
names(pips_complete2)[14] <- 'SpeciesRichness'
str(pips_complete2)

#Grouped boxplot by inoculation and water treatment
ggplot(pips_complete2, aes(x=Inoculation, y=SpeciesRichness, fill=Water)) + 
    geom_boxplot() + theme_light()

#Grouped boxplot by inoculation and plant part
ggplot(pips_complete2, aes(x=Inoculation, y=SpeciesRichness, fill=Water)) + 
    geom_boxplot() + theme_light() + facet_wrap(~ Type)

#write this to .csv
write.csv(pips_complete2, "./combined_treatment_data/pips_complete2.csv")

```

**Outlier analysis: **
```{r}
#load libraries
library(tidyverse)
library(car)
library(ggpubr)
library(effects)
library(emmeans)

dat = read.csv("./combined_treatment_data/pips_complete2_edited.csv") |>
  mutate(
    Inoculation = as.character(Inoculation),
    Water = as.character(Water)
  )

# interaction model
fit1 = aov(log(SpeciesRichness) ~ Type + Inoculation * Water, data = dat)
Anova(fit1, type = "3") # if you have interactions, type-3 sums-of-squares is appropriate

# additive model
fit2 = aov(log(SpeciesRichness) ~ Type + Inoculation + Water, data = dat)
Anova(fit2, type = "2") # if you don't have interactions, type-2 sums-of-squares is appropriate

#plot to check residuals
plot(fit1)
plot(fit2)

# Studentized residuals
MASS::studres(fit1) |>
  abs() |>
  sort()

#remove the outliers, re-run the analysis
filtered_dat = dat[-c(88, 34, 91, 464, 133), ]

# interaction model
fit1 = aov(log(SpeciesRichness) ~ Type + Inoculation * Water, data = filtered_dat)
Anova(fit1, type = "3") # if you have interactions, type-3 sums-of-squares is appropriate

# additive model
fit2 = aov(log(SpeciesRichness) ~ Type + Inoculation + Water, data = filtered_dat)
Anova(fit2, type = "2") # if you don't have interactions, type-2 sums-of-squares is appropriate

#plot to check residuals
plot(fit1)
plot(fit2)

#Re-plot grouped boxplot by inoculation and plant part without the outliers
ggplot(filtered_dat, aes(x=Inoculation, y=log(SpeciesRichness), fill=Water)) + 
    geom_boxplot() + theme_light() + facet_wrap(~ Type)

```

**ANOVA: roots**
```{r}
root_dat = subset(filtered_dat, Type == "Root")

#interaction model
richness_root_fit1 = aov(log(SpeciesRichness) ~ InoculationLabeled * WaterLabeled, data = root_dat)
Anova(richness_root_fit1, type = "3") # if you have interactions, type-3 sums-of-squares is appropriate
plot(richness_root_fit1)
plot(allEffects(richness_root_fit1))

#additive model 
richness_root_fit2 = aov(log(SpeciesRichness) ~ InoculationLabeled + WaterLabeled, data = root_dat)
Anova(richness_root_fit2, type = "2")
plot(richness_root_fit2)
plot(allEffects(richness_root_fit2))
```

**Post-hoc test roots: planned comparisons**
```{r}
## Type III ANOVA ##

#apply the emmeans() command to the ANOVA object.
RichRootsW1<- emmeans(richness_root_fit1, ~ WaterLabeled)
RichRootsW1

#obtain the planned 95% confidence intervals for a pairwise comparison
RootsPlannedW1 <- contrast(RichRootsW1, adjust = "none")
RootsPlannedW1

#apply the emmeans() command to the ANOVA object.
RichRootsI1<- emmeans(richness_root_fit1, ~ InoculationLabeled)
RichRootsI1

#obtain the planned 95% confidence intervals for a pairwise comparison
RootsPlannedI1 <- contrast(RichRootsI1, adjust = "none")
RootsPlannedI1

## Type II ANOVA ##
#apply the emmeans() command to the ANOVA object.
RichRootsW2<- emmeans(richness_root_fit2, ~ WaterLabeled)
RichRootsW2

#obtain the planned 95% confidence intervals for a pairwise comparison
RootsPlannedW2 <- contrast(RichRootsW2, adjust = "none")
RootsPlannedW2

#apply the emmeans() command to the ANOVA object.
RichRootsI2<- emmeans(richness_root_fit2, ~ InoculationLabeled)
RichRootsI2

#obtain the planned 95% confidence intervals for a pairwise comparison
RootsPlannedI2 <- contrast(RichRootsI2, adjust = "none")
RootsPlannedI2
```

**ANOVA: shoots**
```{r}
shoot_dat = subset(filtered_dat, Type == "Shoot")

#interaction model
richness_shoot_fit1 = aov(log(SpeciesRichness) ~ InoculationLabeled * WaterLabeled, data = shoot_dat)
Anova(richness_shoot_fit1, type = "3") # if you have interactions, type-3 sums-of-squares is appropriate
plot(richness_shoot_fit1)
plot(allEffects(richness_shoot_fit1))

#additive model 
richness_shoot_fit2 = aov(log(SpeciesRichness) ~ InoculationLabeled + WaterLabeled, data = shoot_dat)
Anova(richness_shoot_fit2, type = "2")
plot(richness_shoot_fit2)
plot(allEffects(richness_shoot_fit2))
```

**Post-hoc test shoots: planned comparisons**
```{r}
## Type III ANOVA ##

#apply the emmeans() command to the ANOVA object.
RichShootsW1<- emmeans(richness_shoot_fit1, ~ WaterLabeled)
RichShootsW1

#obtain the planned 95% confidence intervals for a pairwise comparison
ShootsPlannedW1 <- contrast(RichShootsW1, adjust = "none")
ShootsPlannedW1

#apply the emmeans() command to the ANOVA object.
RichShootsI1<- emmeans(richness_shoot_fit1, ~ InoculationLabeled)
RichShootsI1

#obtain the planned 95% confidence intervals for a pairwise comparison
ShootsPlannedI1 <- contrast(RichShootsI1, adjust = "none")
ShootsPlannedI1

## Type II ANOVA ##
#apply the emmeans() command to the ANOVA object.
RichShootsW2<- emmeans(richness_shoot_fit2, ~ WaterLabeled)
RichShootsW2

#obtain the planned 95% confidence intervals for a pairwise comparison
ShootsPlannedW2 <- contrast(RichShootsW2, adjust = "none")
ShootsPlannedW2

#apply the emmeans() command to the ANOVA object.
RichShootsI2 <- emmeans(richness_shoot_fit2, ~ InoculationLabeled)
RichShootsI2

#obtain the planned 95% confidence intervals for a pairwise comparison
ShootsPlannedI2 <- contrast(RichShootsI2, adjust = "none")
ShootsPlannedI2

```

**Outlier analysis: below-ground biomass**
```{r}
bio_root = subset(dat, Type == "Root")
  
#interaction model
root_mass1= aov(WeightGrams ~ Inoculation * Water, data = bio_root)
Anova(root_mass1, type = "3") # if you have interactions, type-3 sums-of-squares is appropriate

#additive model
root_mass2 = aov(WeightGrams ~ Inoculation + Water, data = bio_root)
Anova(root_mass2, type = "2")

#plot to check residuals
plot(root_mass1)
plot(root_mass2)

# Studentized residuals
MASS::studres(root_mass1) |>
  abs() |>
  sort()

#remove the outliers, re-run the analysis
filtered_root = bio_root[-c(219), ]

```

**Outlier analysis: above-ground biomass**
```{r}

bio_shoot = subset(dat, Type == "Shoot")

#interaction model
shoot_mass1 = aov(WeightGrams ~ Inoculation * Water, data = bio_shoot)
Anova(shoot_mass1, type = "3") # if you have interactions, type-3 sums-of-squares is appropriate

#additive model
shoot_mass2 = aov(WeightGrams ~ Inoculation + Water, data = bio_shoot)
Anova(shoot_mass2, type = "2")

#plot to check residuals
plot(shoot_mass1)
plot(shoot_mass2)

# Studentized residuals
MASS::studres(shoot_mass1) |>
  abs() |>
  sort()

#remove the outliers, re-run the analysis
filtered_shoot = bio_shoot[-c(384, 439), ]
```

**ANOVA: below-ground biomass**
```{r}
  
#interaction model
root_mass1= aov(WeightGrams ~ InoculationLabeled * WaterLabeled, data = filtered_root)
Anova(root_mass1, type = "3") # if you have interactions, type-3 sums-of-squares is appropriate
plot(root_mass1, main = "Root")
plot(allEffects(root_mass1))

#additive model
root_mass2 = aov(WeightGrams ~ InoculationLabeled + WaterLabeled, data = filtered_root)
Anova(root_mass2, type = "2")
plot(root_mass2, main = "Root")
plot(allEffects(root_mass2))
```

**Post-hoc test: below-ground biomass**
```{r}
## Type III ANOVA ##

#apply the emmeans() command to the ANOVA object.
RootsW1<- emmeans(root_mass1, ~ WaterLabeled)
RootsW1

#obtain the planned 95% confidence intervals for a pairwise comparison
RootsPlannedW1 <- contrast(RootsW1, adjust = "none")
RootsPlannedW1

#apply the emmeans() command to the ANOVA object.
RootsI1<- emmeans(root_mass1, ~ InoculationLabeled)
RootsI1

#obtain the planned 95% confidence intervals for a pairwise comparison
RootsPlannedI1 <- contrast(RootsI1, adjust = "none")
RootsPlannedI1

## Type II ANOVA ##

#apply the emmeans() command to the ANOVA object.
RootsW2<- emmeans(root_mass2, ~ WaterLabeled)
RootsW2

#obtain the planned 95% confidence intervals for a pairwise comparison
RootsPlannedW2 <- contrast(RootsW2, adjust = "none")
RootsPlannedW2

#apply the emmeans() command to the ANOVA object.
RootsI2<- emmeans(root_mass2, ~ InoculationLabeled)
RootsI2

#obtain the planned 95% confidence intervals for a pairwise comparison
RootsPlannedI2 <- contrast(RootsI2, adjust = "none")
RootsPlannedI2
```

**ANOVA: above-ground biomass**
```{r}

#interaction model
shoot_mass1 = aov(WeightGrams ~ InoculationLabeled * WaterLabeled, data = filtered_shoot)
Anova(shoot_mass1, type = "3") # if you have interactions, type-3 sums-of-squares is appropriate
plot(shoot_mass1, main = "Shoot")
plot(allEffects(shoot_mass1))

#additive model
shoot_mass2 = aov(WeightGrams ~ InoculationLabeled + WaterLabeled, data = filtered_shoot)
Anova(shoot_mass2, type = "2")
plot(shoot_mass2, main = "Shoot")
plot(allEffects(shoot_mass2))
```

**Post-hoc test: above-ground biomass**
```{r}
## Type III ANOVA ##

#apply the emmeans() command to the ANOVA object.
ShootsW1<- emmeans(shoot_mass1, ~ WaterLabeled)
ShootsW1

#obtain the planned 95% confidence intervals for a pairwise comparison
ShootsPlannedW1 <- contrast(ShootsW1, adjust = "none")
ShootsPlannedW1

#apply the emmeans() command to the ANOVA object.
ShootsI1<- emmeans(shoot_mass1, ~ InoculationLabeled)
ShootsI1

#obtain the planned 95% confidence intervals for a pairwise comparison
ShootsPlannedI1 <- contrast(ShootsI1, adjust = "none")
ShootsPlannedI1

## Type II ANOVA ##

#apply the emmeans() command to the ANOVA object.
ShootsW2<- emmeans(shoot_mass2, ~ WaterLabeled)
ShootsW2

#obtain the planned 95% confidence intervals for a pairwise comparison
ShootsPlannedW2 <- contrast(ShootsW2, adjust = "none")
ShootsPlannedW2

#apply the emmeans() command to the ANOVA object.
ShootsI2<- emmeans(shoot_mass2, ~ InoculationLabeled)
ShootsI2

#obtain the planned 95% confidence intervals for a pairwise comparison
ShootsPlannedI2 <- contrast(ShootsI2, adjust = "none")
ShootsPlannedI2
```
